---
title: "Time Series Analysis of Climate Data"
output: html_notebook
---

```{r, message=FALSE}
require(mosaic)
options(digits=3)
palette = trellis.par.get()$superpose.symbol$col
library(Stat2Data)
library(forecast)
library(lmtest) # if you want p-values for coefficients in ARIMA models
library(tseries) # need for Augmented Dickey-Fuller test
require(fBasics) # need for normality tests of residuals


```

### Climate Change

The data in **CO2Hawaii** shows the monthly carbon dioxide level (CO2 in ppm) measured at the Mauna Loa Observatory in Hawaii from 1988 through 2017. 

```{r}
data("CO2Hawaii")
head(CO2Hawaii)
tail(CO2Hawaii)
```

We first start by creating a plot of our time series data. See plot below:
```{r}
class(CO2Hawaii)
co2 = ts(CO2Hawaii$CO2)
class(co2)
```

```{r}
plot(co2, ylab = "CO2 levels", xlab = "Time since January 1988", main = "CO2 Levels", col = "blue")
plot(as.ts(co2[324:360]), ylab = "CO2 levels", xlab = "Time since January 2015", main = "CO2 levels over the last 3 years in Hawaii", col = "blue")
```

The graph of the last three years suggests that there is a peak in CO2 levels every year around the 6th month i.e. June and a drop in CO2 levels around October i.e. the 10th month.

























